Portfolios of Subgraph Isomorphism Algorithms

### 2015-12-18 paper/paper.tex
small fixes

\begin{abstract}
Subgraph isomorphism is a computationally challenging problem with important practical
applications, for example in computer vision, biochemistry, and model checking. There are a number
of state-of-the-art algorithms for solving the problem, each of which has its own performance
characteristics. As with many other hard problems, the single best choice of algorithm overall is
rarely the best algorithm on an instance-by-instance. We develop an algorithm
selection approach which leverages
novel features to characterise subgraph isomorphism problems to dynamically decide which algorithm
to use on a per-instance basis. We demonstrate substantial performance improvements on a large set
of hard benchmark problems. In addition, we show how algorithm selection models can be leveraged to
gain new insights into what affects the performance of an algorithm.
\end{abstract}

@@

In Section~\ref{sec:algsel}, we discuss the features that are used to describe
instances, and we describe our algorithm selection approach. It combines a
presolving step, which allows us to easy instances very quickly, with the
algorithm selection approach \LLAMA~\cite{kotthoff_llama_2013}.
In Section~\ref{sec:algsel-exps}, we experimentally evaluate our selection
approach, and show that it is able to close more than \SI{60}{\percent} of the
gap between the single best and the virtual best solver. We conclude and give
directions for future work in Section~\ref{sec:concs}.

@@

In the following, we assume $G_p=(N_p,E_p)$ and $G_t=(N_t,E_t)$ to be the underlying instance of 
subgraph isomorphism.  We  define $n_p = \# N_p$, $n_t = \# N_t$,  $e_p=\# E_p$, $e_t=\#
E_t$, and $d_p$ and $d_t$ to be the maximum degrees of the graphs $G_p$ and $G_t$.

@@

We have selected a set of algorithms with complementary performance. These
algorithms will compose our portfolio in the per-instance selection approach
described in Section~\ref{sec:algsel}.
\begin{itemize}
\item \VFtwo{}~\cite{Cordella:2004} performs weak filtering that is especially fast on
    trivially satisfiable instances;
\item \LAD{}~\cite{Solnon:2010} combines two strong but expensive filtering techniques
    (All-Different Generalized Arc Consistency and Locally All-Different);
\item \Glasgow{}~\cite{McCreesh:2015} does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic All-Different propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

@@

We introduce two new variants of \LAD. The first, called \IncompleteLAD, does weaker
filtering which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check that there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
\IncompleteLAD is an incomplete algorithm that checks a sufficient, but not
necessary, condition for inconsistency: when it does not detect inconsistency,
the instance may still be unsatisfiable. Its main benefit is that it runs very
fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

@@

\noindent Note that Classes 3 and 12 were not considered in the previous
experimental study by McCreesh and Prosser~\cite{McCreesh:2015}.  This set of
instances is also much larger than that of Battiti and
Mascia~\cite{battiti-mascia07}, who were the first to propose algorithm
portfolios for subgraph isomorphism problems.  Battiti and Mascia only
considered a pure parallel portfolio consisting of two randomised solvers, and
without a selection mechanism. Their problem set consisted entirely of
satisfiable instances.

@@

Figure~\ref{expTimeGraph} displays the evolution of the cumulative number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. \IncompleteLAD is
able to solve easy unsatisfiable instances very quickly, in a few milliseconds. Hence, for time
limits lower than \SI{5}{\ms}, the best solver is \IncompleteLAD. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satisfiable instances.

\PathLAD and \GlasgowOne outperform \IncompleteLAD for longer time limits: \PathLAD is the
best solver for time limits larger than \SI{5}{\ms} and lower than \SI{40}{\ms}, and \GlasgowOne is the best solver
for time limits larger than \SI{40}{\ms} and lower than \SI{3000}{\ms}.

\GlasgowTwo becomes the best solver for time limits larger than \SI{3000}{\ms}.
As we increase the CPU time limit, the performance of variants of \Glasgow with
longer paths (\GlasgowThree and \GlasgowFour) improves. This is what we expect,
as more reasoning is expensive, but increases the potential reduction of the
search space. Eventually, \GlasgowTwo and \GlasgowThree become very closely
matched, and with runtimes very close to the limit, \GlasgowFour nearly catches
up. (This behaviour is class-dependent: over class 2, for example, the behavior
is roughly monotone, with \GlasgowOne dominating for low runtimes, then
\GlasgowTwo, then \GlasgowThree, then \GlasgowFour each becoming best as the runtimes
increase.)

@@

Our approach is composed of three steps. First, we run two presolvers in a static way to quickly
solve easy instances; second, we extract features from instances which are not solved by the first
step; and third, we select an algorithm and run it.

@@

Experimental results reported in Section~\ref{expComp} have shown us that \IncompleteLAD is very fast
(\SI{7}{\ms} on average) and is able to solve 1919 instances from our benchmark
set very quickly. Therefore, we first run
\IncompleteLAD: if unsatisfiability is detected, we do not need to process it further.

\VFtwo is also able to solve many easy instances very quickly: among the 3806 instances which are not
solved by \IncompleteLAD, 1470 are solved by \VFtwo in less than 50 milliseconds. Therefore, after
running \IncompleteLAD, we run the \VFtwo solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.

We include \VFtwo in the portfolio, as it may solve an instance given more time,
but not \IncompleteLAD.

@@

If presolving does not give us a solution, we extract features. For both the pattern and
the target graph, we consider some basic graph properties, which can be computed
very quickly:

\begin{itemize}
    \item the number of vertices and edges;
    \item the density---we expect that some kinds of filtering (like those based upon locally
        all-different constraints) might be expensive and ineffective on
        dense graphs;
    \item how many loops (self-adjacent vertices) the graph contains---since loops must be mapped to
        loops, this could have a strong effect on how easy an instance is;
    \item the mean and maximum degrees, and whether or not every vertex has the same degree (the
        degree-based invariants used by \LAD and \Glasgow do nothing at the top of search if every
        vertex has the same degree);
    \item whether or not the graph is connected;
    \item the mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective) and
        the proportion of vertex pairs which are at least 2, 3 and 4 apart.
\end{itemize}

\noindent Alongside these basic features, we include information computed by \IncompleteLAD. To (try to) prove
inconsistency, \IncompleteLAD removes candidate pairs. The number of successfully removed pairs
gives information on the distribution of edges (the fewer removed pairs, the more uniform the
distribution). As well as the number of removed pairs, we also record the percentage with respect to
all possible pairs, and the minimum and maximum percentages of removed values on
a per-variable basis.

@@

Table~\ref{tab:res} shows the performance of our algorithm selection approach,
compared to two baselines, on the reduced set of 2336 instances. The virtual
best solver is the oracle predictor that, for each instance, chooses the best
solver from our portfolio. This is the upper bound of what an algorithm
selection approach can achieve. The single best solver is the one solver from
the portfolio that has the overall best performance across the entire set of
instances, at the CPU time limit of \SI{e8}{\ms}, i.e.\ \GlasgowTwo. We consider
it a lower bound on the performance of the algorithm selection approach.  We are
able to solve 30 more instances than the single best solver within the timeout,
with only an additional 16 to the virtual best. In terms of average performance,
we are able to close \SI{64}{\percent} of the gap between the single best and
the virtual best solver.

@@

\begin{table}[p]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 2219 & 5822809\\
\LLAMA & 705097 & 2203 & 6529563\\
\GlasgowTwo & 1960683 & 2173 & 7783492\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection performance on the reduced set of 2336 instances. MCP is the
misclassification penalty; that is, the additional time required to solve an instance because of
choosing solvers that perform worse than the best. Mean MCP and performance are over all 2336 instances; when an
instance is not solved, its performance is set to the time limit
(\SI{e8}{\ms}).}\label{tab:res}
\end{table}

@@

\caption{Cumulative number of solved instances over time for the virtual best solver,
\LLAMA, and the single best solver \GlasgowTwo on the reduced set of 2336
instances. Other individual solvers are shown as dotted lines, in the same
colours as Figure~\ref{expTimeGraph}. }
\label{fig:portfolio-ecdf}
\end{figure}

Table~\ref{tab:res} shows the performance of the selection model on its own. The
performance of the entire algorithm selection system, including the
preprocessing, is shown in Table~\ref{tab:resfull}. Our system is able to close
more than \SI{60}{\percent} of the gap between single and virtual best, similar
to the results on the reduced set of instances.

@@

\begin{figure}[p]
    \input{graph-full}
\caption{Cumulative number of solved instances over time for the virtual best solver,
\LLAMA, and the single best solver \GlasgowTwo on the full set of 5725
instances. Other individual solvers are shown as dotted lines, in the same
colours as Figure~\ref{expTimeGraph}. }
\label{fig:portfolio-ecdf-full}

@@

Even though Figures~\ref{fig:portfolio-ecdf} and~\ref{fig:portfolio-ecdf-full}
show that on easy instances, the model does not perform well, it does over
the entire set of instances (cf.\ Tables~\ref{tab:res} and~\ref{tab:resfull}).

@@

As expected, both the pattern graph and the target graph provide important
features. We also conclude that even basic graph properties are predictive of
sophisticated algorithms' performance.

@@

To gain further insight into the behaviour of the algorithms, we
investigated what affects the relative performance of \PathLAD and \GlasgowTwo.
This pair is of particular interest because they are the best ``medium-case''
algorithms that use strong and weak filtering during search respectively. We
used machine learning techniques to train a simple, human-understandable model
which is able to distinguish these solvers for the 2336 hard instances and gives
performance better than always choosing one of them. The model uses four rules:

@@

This provides an interesting insight into the behaviour of our algorithms, as
well as giving indications for future work.

@@

The problem of identifying subgraph isomorphisms is a hard computation problem that has many
applications in diverse areas. In this paper, we presented a portfolio of six algorithms from the
literature and two new variants of the \LAD algorithm. We introduced a set of novel features to
characterise subgraph isomorphism problems and leveraged them to select the most appropriate
algorithm from the portfolio for each instance.

>>>>

\begin{abstract}
Subgraph isomorphism is a computationally challenging problem with important practical
applications, for example in computer vision, biochemistry, and model checking. There are a number
of state-of-the-art algorithms for solving the problem, each of which has its own performance
characteristics. As with many other hard problems, the single best choice of algorithm overall is
rarely the best algorithm on an instance-by-instance. We develop an algorithm
selection approach which leverages
novel features to characterise subgraph isomorphism problems and dynamically
decides which algorithm
to use on a per-instance basis. We demonstrate substantial performance improvements on a large set
of hard benchmark problems. In addition, we show how algorithm selection models can be leveraged to
gain new insights into what affects the performance of an algorithm.
\end{abstract}

@@

In Section~\ref{sec:algsel}, we discuss the features that are used to describe
instances, and we describe our algorithm selection approach. It combines a
presolving step, which allows us to easy instances very quickly, with an
algorithm selection step that uses \LLAMA~\cite{kotthoff_llama_2013}.
In Section~\ref{sec:algsel-exps}, we experimentally evaluate our selection
approach and show that it is able to close more than \SI{60}{\percent} of the
gap between the single best and the virtual best solver. We conclude and give
directions for future work in Section~\ref{sec:concs}.

@@

We define $n_p = \# N_p$, $n_t = \# N_t$,  $e_p=\# E_p$, $e_t=\# E_t$, and $d_p$
and $d_t$ to be the maximum degrees of the graphs $G_p$ and $G_t$.

@@

We selected the following algorithms from the literature, based on their
performance.
\begin{itemize}
\item \VFtwo{}~\cite{Cordella:2004} performs weak filtering that is especially fast on
    trivially satisfiable instances;
\item \LAD{}~\cite{Solnon:2010} combines two strong but expensive filtering techniques
    (All-Different Generalized Arc Consistency and Locally All-Different);
\item \Glasgow{}~\cite{McCreesh:2015} does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic All-Different propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

@@

We introduce two new variants of \LAD. The first, called \IncompleteLAD, does weaker
filtering which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check
if there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
\IncompleteLAD is an incomplete algorithm that checks a sufficient, but not
necessary, condition for inconsistency: when it does not detect inconsistency,
the instance may still be unsatisfiable. Its main benefit is that it runs very
fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

@@

\noindent Note that Classes 3 and 12 were not considered in the previous
experimental study by McCreesh and Prosser~\cite{McCreesh:2015}.  Our set of
instances is much larger than that of Battiti and
Mascia~\cite{battiti-mascia07}, who were the first to propose algorithm
portfolios for subgraph isomorphism problems.  Battiti and Mascia only
considered a pure parallel portfolio consisting of two randomised solvers
without a selection mechanism. Their problem set consisted entirely of
satisfiable instances.

@@

Figure~\ref{expTimeGraph} displays the evolution of the cumulative number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. \IncompleteLAD is
able to solve easy unsatisfiable instances very quickly, in a few milliseconds.
For time
limits less than \SI{5}{\ms}, it is the best solver. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satisfiable instances.

\PathLAD and \GlasgowOne outperform \IncompleteLAD for longer time limits: \PathLAD is the
best solver for time limits greater than \SI{5}{\ms} and less than \SI{40}{\ms}, and \GlasgowOne is the best solver
for time limits greater than \SI{40}{\ms} and less than \SI{3000}{\ms}.

\GlasgowTwo becomes the best solver for time limits greater than \SI{3000}{\ms}.
As we increase the CPU time limit, the performance of variants of \Glasgow with
longer paths (\GlasgowThree and \GlasgowFour) improves. This is what we expect,
as more reasoning is expensive, but increases the potential reduction of the
search space. Eventually, \GlasgowTwo and \GlasgowThree become very closely
matched, and with runtimes very close to the limit, \GlasgowFour nearly catches
up. This behavior is class-dependent: for class 2, for example, the behavior
is roughly monotone, with \GlasgowOne dominating for low runtimes, then
\GlasgowTwo, then \GlasgowThree, then \GlasgowFour each becoming best as the runtimes
increase.

@@

Our approach is composed of three steps. First, we run two presolvers in a
static way to quickly solve easy instances. This ensures that we achieve good
performance on instances that can be solved in a small amount of time. Second,
we extract features from instances which are not solved by the first step.
Finally, we run algorithm selection to choose the algorithm to solve the
instance with.

@@

Experimental results reported in Section~\ref{expComp} show that \IncompleteLAD is very fast
(\SI{7}{\ms} on average) and able to solve 1919 instances from our benchmark
set very quickly. Therefore, we first run
\IncompleteLAD: if unsatisfiability is detected, we do not need to process it further.

\VFtwo is also able to solve many easy instances very quickly: from the 3806
instances that are not
solved by \IncompleteLAD, 1470 are solved by \VFtwo in less than 50 milliseconds. Therefore, after
running \IncompleteLAD, we run the \VFtwo solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.

We also include \VFtwo in the portfolio, as it may solve an instance given more
time, but not \IncompleteLAD, as it is an incomplete solver that cannot solve
satisfiable instances.

@@

If presolving does not give us a solution, we extract features that characterize
the instances. For both the pattern and
the target graph, we consider some basic graph properties that can be computed
very quickly:

\begin{itemize}
    \item the number of vertices and edges;
    \item the density---we expect that some kinds of filtering (like those based upon locally
        all-different constraints) might be expensive and ineffective on
        dense graphs;
    \item how many loops (self-adjacent vertices) the graph contains---as loops must be mapped to
        loops, this could have a strong effect on how easy an instance is;
    \item the mean and maximum degrees, and whether or not every vertex has the same degree (the
        degree-based invariants used by \LAD and \Glasgow do nothing at the top of search if every
        vertex has the same degree);
    \item whether or not the graph is connected;
    \item the mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective) and
        the proportion of vertex pairs which are at least 2, 3 and 4 apart.
\end{itemize}

\noindent Alongside these basic features, we include information computed by \IncompleteLAD. To (try to) prove
inconsistency, \IncompleteLAD removes candidate pairs. The number of successfully removed pairs
gives information on the distribution of edges (the fewer removed pairs, the more uniform the
distribution). As well as the number of removed pairs, we also record the percentage with respect to
all possible pairs, and the minimum and maximum percentages of removed values on
a per-variable basis. Finally, we include the CPU time required to compute these
features as features.

@@

Table~\ref{tab:res} shows the performance of our algorithm selection approach,
compared to two baselines, on the set of 2336 hard instances. The virtual
best solver is the oracle predictor that, for each instance, chooses the best
solver from our portfolio. This is the upper bound of what an algorithm
selection approach can achieve. The single best solver is the one solver from
the portfolio that has the overall best performance across the entire set of
instances, at the CPU time limit of \SI{e8}{\ms}, i.e.\ \GlasgowTwo. We consider
it a lower bound on the performance of the algorithm selection approach.  We are
able to solve 30 more instances than the single best solver within the timeout,
with only an additional 16 to the virtual best. In terms of average performance,
we are able to close \SI{64}{\percent} of the gap between the single best and
the virtual best solver.

@@

\begin{table}[p]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 2219 & 5822809\\
\LLAMA & 705097 & 2203 & 6529563\\
\GlasgowTwo & 1960683 & 2173 & 7783492\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection performance on the set of 2336 hard instances. MCP is the
misclassification penalty; that is, the additional time required to solve an instance because of
choosing solvers that perform worse than the best. Mean MCP and performance are over all 2336 instances; when an
instance is not solved, its performance is set to the time limit
(\SI{e8}{\ms}).}\label{tab:res}
\end{table}

@@

\caption{Cumulative number of solved instances over time for the virtual best solver,
\LLAMA, and the single best solver \GlasgowTwo on the set of 2336 hard
instances. Other individual solvers are shown as dotted lines, in the same
colors as Figure~\ref{expTimeGraph}. }
\label{fig:portfolio-ecdf}
\end{figure}

Table~\ref{tab:res} shows the performance of the selection model on its own. The
performance of the entire algorithm selection system, including the
preprocessing, is shown in Table~\ref{tab:resfull}. Our system is able to close
more than \SI{60}{\percent} of the gap between single and virtual best, similar
to the results on the set of hard instances.

@@

\begin{figure}[p]
    \input{graph-full}
\caption{Cumulative number of solved instances over time for the virtual best solver,
\LLAMA, and the single best solver \GlasgowTwo on the full set of 5725
instances. Other individual solvers are shown as dotted lines, in the same
colors as Figure~\ref{expTimeGraph}. }
\label{fig:portfolio-ecdf-full}

@@

Figures~\ref{fig:portfolio-ecdf} and~\ref{fig:portfolio-ecdf-full}
show that for the easy instances from the set of hard instances, the performance
improvement through algorithm selection is negated by the cost of computing the
features. The presolving steps improve performance dramatically over the full
set of instances (cf.\ Tables~\ref{tab:res} and~\ref{tab:resfull}).

@@

As expected, both the pattern graph and the target graph provide important
features. We conclude that even basic graph properties are predictive of
sophisticated algorithms' performance.

@@

To gain further insight into the behavior of the algorithms, we
investigated what affects the relative performance of \PathLAD and \GlasgowTwo.
This pair is of particular interest because they are the best ``medium-case''
algorithms that use strong and weak filtering during search, respectively. We
used machine learning techniques (JRip~\cite{Cohen1995}) to train a simple,
human-understandable model which is able to distinguish these solvers for the
2336 hard instances and gives performance better than always choosing one of
them. The model uses four rules:

@@

This provides an interesting insight into the behavior of our algorithms, as
well as giving indications for future work.

@@

The problem of identifying subgraph isomorphisms is a hard computational problem that has many
applications in diverse areas. In this paper, we presented a portfolio of six algorithms from the
literature and two new variants of the \LAD algorithm. We introduced a set of novel features to
characterise subgraph isomorphism problems and leveraged them to select the most appropriate
algorithm from the portfolio for each instance.
